Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-04 2024pipit/section/20-Motivation.tex
kill scratch

\subsection{Scratch}

For the trigger fetch logic, we will use count_when as the index into the triggers array.
To retrieve the element from the triggers array, we assume a pure, non-streaming \fstar{} function for array indexing.
We use an unsafe indexing function, which we lift to a streaming primitive and wrap in a contract to make it safe:

@@

Contracts consist of a property status, a rely clause, a body, and a guarantee clause.
The rely clause states that the calling context must ensure that every element of the index stream is a valid index.
In turn, the contract guarantees that every element of the result will be the element of the array at the corresponding index.
The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
This is a somewhat pathological contract, as the guarantee does not abstract away the details of the implementation at all.

We use the unsafe indexing function, rather than a safe variant with a precondition, as we cannot currently lift primitives with preconditions; we discuss future work to remove this limitation in \autoref{s:core:primitive-tables}.

We can now combine the count_when and index stream functions to implement a (much simplified) time-triggered trigger fetch controller.

\begin{tabbing}
  @MM@\= @MM@ \= @let index@ \= \kill
  @let@ trigger_fetch ($\textit{time}$: stream $\NN$): stream trigger = \\
    \> $@rec @ \textit{trigger}.$ \\
    \> \> $@let @ \textit{inc} = \text{false} @ fby @ (\textit{trigger.time_mark} < \textit{time}) @ in@$\\
    \> \> $@let @ \textit{index} = \mbox{count_when}_{\PSValid}~(\text{Array.size}~\textit{trigger_array} - 1)~\textit{inc} @ in@$ \\
    \> \> $\text{array_index}~\textit{trigger_array}~\textit{index}$\\
\end{tabbing}

The trigger_fetch function takes a stream denoting the current time-since-reset in microseconds and returns a stream of the current active trigger.
Each trigger also has a time-mark, which denotes when it should start.
To compute the index, we use the count_when function, so that it increments when the \emph{previous} trigger's time is in the past.
We then use the array_index function to look up the new trigger in the constant array of triggers.

Finally, we need to prove that trigger_fetch's contract instantiation is valid.
This proof script is much the same as the previous one:
\begin{tabbing}
  @MM@\= @MM@ \= \kill
  @let@ $\mbox{trigger_fetch}_{\PSValid}$ ($\textit{max}$: $\NN$): stream $\NN$ $\to$ stream trigger = \\
    \> @let@ $\textit{system} = \mbox{System.translate}_1~\text{trigger_fetch}$ @in@ \\
    \> @assert@ (System.inductive_check $\textit{system}$) @by@ (pipit\_simplify ()); \\
    \> $@bless@_1$ trigger_fetch
\end{tabbing}

Since the implementation of trigger_fetch used $\text{count_valid}_{\PSValid}$ instead of count_valid, the proof did not need to re-prove the property about the range of count.

We have only proved a simple indexing property here but have seen the important pieces of checks, contracts and deferred proofs.
In \autoref{s:evaluation} we will extend this small example to the real system, which uses checked machine-integer arithmetic and proves that scheduling is never late.

>>>>

@@

Contracts consist of a property status, a rely clause, a body, and a guarantee clause.
The rely clause states that the calling context must ensure that every element of the index stream is a valid index.
In turn, the contract guarantees that every element of the result will be the element of the array at the corresponding index.
The contract status of unknown ($\PSUnknown$) denotes that, although the body satisfies the contract, we have not yet shown that the calling context satisfies the rely clause -- such a proof must wait until the contract is instantiated.
